Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-23 icfp.tex
3.5 and start 3.6

% Recursive let bindings are very similar to non-recursive ones, the main
% exception being that the recursive bindings may be defined in terms of
% themselves, and we may have more than one binding.
%
Recursive let bindings also bind expressions \emph{lazily}, so, similarly, we
introduce a $\D$-variable for each binding. Resources required to type the
let-bindings are still available in the body of the let, to later be consumed
via $\D$-variables, or, directly, if the let-bindings are unused.
%
Additionally, as seen in Section
\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
binder in its own definition entail consuming all resources otherwise required
to type the binder's body.

As for groups of recursive let bindings, we recall that, in our
system, all recursive let groups are strongly-connected -- all bindings in a
recursive let group are mutually recursive in the sense that they all
(transitively) depend on one another.
%
% In light of it, (mutually) recursive uses of binders entail consuming all the
% resources required to type all of the binders.
%
By definition of strongly-connected, any binder in the mutually recursive group
may end up using any binder from the group. Accordingly, occurrences of any
such binder must be typed conservatively, using up the sum of resources needed
to type each of the binders,
%
i.e., all binders in a strongly-connected group of mutually recursive let
bindings have to be typed with the same linear resources (which are the least
upper bound of resources needed to type the bodies of all binders, accounting
for uses of mutually-recursive binders).
%
% Therefore, we conclude the least upper bound of resources required to type a
% mutually recursive group of let bindings to be the same for any such group that
% is strongly-connected.

The typing rule for recursive groups of bindings leverages both the assumption
that all recursive let bindings are strongly connected and the observation that
every binder in a strongly connected group of recursive bindings is typed with
the same linear context. Consequently, all bindings of the recursive group are
introduced as $\D$-vars with the same $\D$ environment -- using any one of the
bindings in a recursive group entails consuming all resources required to type
that same group, which is also why we can use the same linear resources to type
each binder:
\[
\TypeLetRec
\]
This formulation is not syntax-directed 
% (from which an implementation is direct) 
because determining a particular $\D$
to type and annotate all binders is not obvious. We present our system and
metatheory agnostically to the challenge of inferring this linear typing
environment by assuming recursive let expressions are annotated with the
correct typing environment.

@@

Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
a case expression \emph{evaluates its scrutinee} to Weak Head Normal
Form (WHNF)~\cite{10.5555/1096899}, and \emph{then} selects the case
alternative corresponding to the pattern matching the WHNF of the
scrutinee\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}. An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $\lambda x.~e$ 
or a datatype constructor application $K~\ov{e}$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF. % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed (these resources will be consumed when the expressions are evaluated).

>>>>

% Recursive let bindings are very similar to non-recursive ones, the main
% exception being that the recursive bindings may be defined in terms of
% themselves, and we may have more than one binding.
%
Recursive let bindings also bind expressions lazily, so, similarly, we
introduce a $\D$-variable for each binding. Resources required to type the
let-bindings are still available in the body of the let, to later be consumed
via $\D$-variables, or, directly, if the let-bindings are unused.
%
%Additionally, as seen in Section
%\ref{sec:semantic-linearity-examples:recursive-lets}, recursive uses of a
%binder in its own definition entail consuming all resources otherwise required
%to type the binder's body.

%We assume all mutually recursive groups of bindings are strongly-connected.
%As for groups of recursive let bindings, we recall that, in our
%system, all recursive let groups are strongly-connected -- all bindings in a
%recursive let group are mutually recursive in the sense that they all
%(transitively) depend on one another.
%
% In light of it, (mutually) recursive uses of binders entail consuming all the
% resources required to type all of the binders.
%
%By definition, any binder in such a group may end up recursively using
%any other binder from the group and thus use up any resource occurring.
%
%Accordingly, occurrences of any such binder
%must be typed conservatively, using up the sum of resources needed to type each
%of the binders,
%%
%i.e.,
%Accordingly, all the binders in a strongly-connected group of mutually recursive let
%bindings have to be typed with the same linear resources (which are the least
%upper bound of resources needed to type the bodies of all binders, accounting
%for uses of mutually-recursive binders).
%
% Therefore, we conclude the least upper bound of resources required to type a
% mutually recursive group of let bindings to be the same for any such group that
% is strongly-connected.

The typing rule for recursive groups of bindings leverages both the assumption
that all mutually recursive groups are strongly connected and the corollary
that every binder in such a group must be typed with the same linear context,
as observed in \ref{sec:semantic-linearity-examples:recursive-lets}.
Consequently, all bindings of the recursive group are introduced as $\D$-vars
which all share the same $\D$ environment -- using any one of the bindings in a
recursive group entails consuming all resources required to type that same
group -- so, at most a single binder from the group can be used in the let-continuation.
%\todo{should we say how this means at most once of the binders from the group can be used?} OK
%, which is also why we can use the same linear resources to type each binder:
\[
\TypeLetRec
\]
This formulation is not syntax-directed 
% (from which an implementation is direct) 
because determining a particular $\D$
to type and annotate all binders is not obvious. We present our system and
metatheory agnostically to the challenge of inferring this linear typing
environment by assuming recursive let expressions are annotated with the
correct typing environment.

@@

%Case expressions \emph{drive evaluation} --
% thus they are key to realize a type system that understands linearity in the presence of lazy evaluation.
%
%However, the evaluation of case expressions is considerably nuanced,
%
%In lazy let bindings, computations  can be a case
%expression can effectively consume resources rather than just 
%
A case expressions drive evaluation by evaluating their scrutinee to Weak Head
Normal Form (WHNF)~\cite{10.5555/1096899}. Then, the case
alternative whose pattern matches the result of evaluating the
scrutinee is taken\footnote{In our calculus, the alternatives are assumed to be
  exhaustive, i.e. there always exists at least one pattern which
  matches the scrutinee in its WHNF, so we're guaranteed to have an
  expression to progress evaluation.}.
An expression in WHNF can either be %\begin{itemize}
% \item
a $\lambda$-abstraction $(\lambda x.~e)$ 
or a datatype constructor application $(K~\ov{e})$.
%\end{itemize}
In both cases, the sub-expressions $e$ or $\ov{e}$ occurring in the lambda body
or as constructor arguments need not be evaluated for the lambda or constructor
application to be in WHNF. % (if otherwise all sub-expressions
% were fully evaluated the whole expression would also be in \emph{normal form}).
% (that is why it is called \emph{weak} head normal form).
%
Accordingly, these sub-expressions might still depend on linear resources to be
well-typed (such resources will only be consumed when the sub-expressions themselves are evaluated).
